# Early Bronze Remains {#earlybronze}

## Preservation and Demography

Generally, preservation of the sample is fair, although there is differential preservation between and within tombs. Remains from Tombs 1 and 4 exhibit the best overall preservation. Individuals with poor preservation were generally incomplete, highly fragmentary, or the bones had undergone significant taphonomic degradation.

For each tomb, the minimum number of individuals (MNI) was determined by considering repeated skeletal elements and matching commingled remains based upon diagnostic indicators for age and sex, as well as general appearance. In some instances, especially if the remains were highly fragmentary, a large portion of the remains could not be assigned to any particular individual (e.g., Tombs 3 and 7). Human remains recovered from the tomb complex (up to the 2006 field season) represent an MNI of 35 individuals (see Table 1).

## Paleopathology

### General Observations

Due to diff erential preservation, paleopathological data are not reported for nearly one-third of the sample. Also, because of time constraints, paleopathology was not recorded for the individuals from Tomb 8. Of those observed, 13/21 individuals exhibit some type of pathological lesion.

The most common pathology is periostitis, which is commonly found in archaeological cemetery samples [@ortner2003identification]. Periostitis occurs in 6/21 of the observable sample and more frequently on the lower limbs, although the upper limbs are also aff ected. Both active and healed lesions were observed. Porotic hyperostosis, another lesion commonly reported in archaeological skeletal samples, is uncommon in the Umm el-Marra sample. Porotic hyperostosis is often considered a result of anemia, although its etiology is likely more complicated [@walker2009causes]. No cases of porotic hyperostosis on the cranial vault occur; however, cribra orbitalia---porosities on the orbital roof---occur in 2/21 individuals.

### Selected Cases


## Dietary Reconstruction

### Stable Isotope Analysis

### Dental Wear


## Dental Nonmetrics

Dental nonmectric data were collected using the UADAS recording system [@scott2000anthteeth]. Population distance was measured using Smith's Mean Measure of Divergence (MMD). @soltysiak2011mmd provides an R script for calculating the MMD.

```{r echo=FALSE, message=FALSE, results='hide'}
## Read in data
M_n <- read.csv("https://www.dropbox.com/s/6a2bkv8yw42unlx/nonmetric-dental_obs2.csv?raw=1", header = TRUE)
# M_p <- read.csv("https://www.dropbox.com/s/0cm16evzsbyt5dh/nonmetric-dental_prop2.csv?raw=1", header = TRUE)
M_p <- read.csv("https://www.dropbox.com/s/ekgox649bprmvgs/nonmetric-dental_prop3.csv?raw=1", header = TRUE)  ## With Bartlett correction


## Read in data. The files below include larger set of nonmetric traits
#M_n <- read.csv("https://www.dropbox.com/s/svynefmmzrdjirz/nonmetric-dental_obs.csv?raw=1", header = TRUE)
#M_p <- read.csv("https://www.dropbox.com/s/alwnmdawch6afyi/nonmetric-dental_prop.csv?raw=1", header = TRUE)


## Anscombe transformation
theta <- function(n,p) { asin((n/(n+3/4))*(1-2*p)) }
## Freeman & Tukey transformation
#theta <- function(n,p) { 0.5*(asin(1-(2*p*n/(n+1)))+asin(1-(2*((p*n)+1)/(n+1)))) }

## Corrections for small sample size
## Freeman & Tukey correction
thetadiff <- function(nA,pA,nB,pB) { (theta(nA,pA) - theta(nB,pB))^2 - (1/(nA+0.5) + 1/(nB+0.5)) }
## Grewal correction
#thetadiff <- function(nA,pA,nB,pB) { (theta(nA,pA) - theta(nB,pB))^2 - (1/nA + 1/nB) }
## uncorrected formula
#thetadiff <- function(nA,pA,nB,pB) { (theta(nA,pA) - theta(nB,pB))^2 }

## -------SECTION C: VARIABLE STATUS-----------

VarMatrix <- M_n[1:2,2:length(M_n[1,])]
MMDMatrix <- matrix(0,length(M_n[,1]),length(M_n[,1]))

for (a in seq_along(VarMatrix[1,])) {
for (b in seq_along(MMDMatrix[,1])) {
for (c in seq_along(MMDMatrix[1,])) {
## percent frequencies
#MMDMatrix[b,c] <- thetadiff(M_n[b,a+1], M_p[b,a+1]/100, M_n[c,a+1], M_p[c,a+1]/100)
## proportions
MMDMatrix[b,c] <- thetadiff(M_n[b,a+1], M_p[b,a+1], M_n[c,a+1], M_p[c,a+1])
} }

for (b in seq_along(MMDMatrix[,1])) {
for (c in seq_along(MMDMatrix[1,])) {
if (b >= c) { MMDMatrix[b,c] = 0 } } }

VNeg <- 0
VPos <- 0
for (b in seq_along(MMDMatrix[,1])) {
for (c in seq_along(MMDMatrix[1,])) {
if (MMDMatrix[b,c] > 0) { VPos = VPos + 1 }
if (MMDMatrix[b,c] < 0) { VNeg = VNeg + 1 } } }

VarMatrix[1,a] = sum(MMDMatrix)
VarMatrix[2,a] = VPos / (VPos + VNeg) }

VarStatus <- t(VarMatrix)

## --------SECTION D: MMD MATRIX---------------
MMDMatrix <- matrix(0,length(M_n[,1]),length(M_n[,1]))
dimnames(MMDMatrix) <- list(M_n[,1], M_n[,1])

for (a in seq_along(MMDMatrix[,1])) {
for (b in seq_along(MMDMatrix[1,])) {
MMDVect <- vector("double",length(M_n[1,])-1)
## percent frequencies
#for (i in seq_along(MMDVect)) { MMDVect[i] <- thetadiff(M_n[a,i+1], M_p[a,i+1]/100, M_n[b,i+1], M_p[b,i+1]/100) }
## proportions
for (i in seq_along(MMDVect)) { MMDVect[i] <- thetadiff(M_n[a,i+1], M_p[a,i+1], M_n[b,i+1], M_p[b,i+1]) }
MMDMatrix[a,b] <- sum(MMDVect) / length(MMDVect) } }

## forced 0 when a sample is compared to itself
for (a in seq_along(MMDMatrix[,1])) { MMDMatrix[a,a] = 0 }

## -------------SECTION E: SD MATRIX-------------------------------------------

## standard deviation for MMD, Sjovold's formula
SDMatrix <- matrix(0,length(M_n[,1]),length(M_n[,1]))
dimnames(SDMatrix) <- list(M_n[,1], M_n[,1])
SDDiff <- function(nA,nB) { (1/nA + 1/nB)^2 }

for (a in seq_along(MMDMatrix[,1])) {
for (b in seq_along(MMDMatrix[1,])) {
SDVect <- vector("double",length(M_n[1,])-1)
for (i in seq_along(SDVect)) { SDVect[i] <- SDDiff(M_n[a,i+1], M_n[b,i+1]) }
SDMatrix[a,b] <- sqrt(sum(SDVect)*2 / length(SDVect)^2) } }

## --------SECTION F: SIGNIFICANCE MATRIX---------------

## statistical significance
SigMatrix <- matrix(1,length(M_n[,1]),length(M_n[,1]))
dimnames(SigMatrix) <- list(M_n[,1], M_n[,1])

for (a in seq_along(MMDMatrix[,1])) {
for (b in seq_along(MMDMatrix[1,])) {

dist <- MMDMatrix[a,b] / SDMatrix[a,b]
SigMatrix[a,b] = round((1-pnorm(dist))*2, digits = 8)
if (MMDMatrix[a,b] < 0) SigMatrix[a,b] = 1
} }

## ---------SECTION G: OUTPUT-------------

#VarStatus
write.csv(VarStatus, file="output_vars.csv")

#MMDMatrix
write.csv(MMDMatrix, file="output_mmd.csv")

#SDMatrix
write.csv(SDMatrix, file="output_mmd-sd.csv")

#SigMatrix
write.csv(SigMatrix, file="output_mmd-p.csv")

```

```{r, echo=FALSE}
knitr::kable(MMDMatrix, pad = 0.01, caption = 'MMD Matrix for Umm el-Marra and Comparative Samples')
```

Also of interest is the significance matrix.

```{r, echo=FALSE}
knitr::kable(SigMatrix, caption = 'Significance Matrix for Umm el-Marra and Comparative Samples')
```

Also of interest is the standard deviation matrix.

```{r, echo=FALSE}
knitr::kable(SDMatrix, caption = 'Standard Deviation Matrix for Umm el-Marra and Comparative Samples')
```
